Characterization of the microbial and fungal communities in ancient pottery from Neolithic southwestern China

### 2024-01-30 analysis/paper/paper.qmd
work on cultural background and method

# Introduction
Ceramics are the most common artifacts unearthed from archaeological sites, offering valuable insights into diverse anthropological facets encompassing cultural, social, and economic dimensions. Foodway is one of the topics that can can explored based on food residues absorbed into the matrices of ceramics. Food residues are useful sources to explore pottery use and past culinary practices. This analytical pursuit heavily relies on contemporary scientific methodologies. For example, organic residues, particularly lipids, extracted from pottery through a series of chemical techniques enable the identification of terrestrial animals, marine resources, and plants [@Evershed2008; @Evershed2002; @Evershed2008marine; @Craig2011; @Heron2016]. Furthermore, the identification of foodstuffs contained in storage vessels can be achieved through ancient DNA analysis, providing direct evidence of pottery content at the genes or species levels [@Foley2012; @Hansson2008; @Robinson2017]. These analytical approaches underscore the significance of preserving the condition of food residues within archaeological contexts, ensuring the efficacy of these scientific methodologies.

@@

In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery through a comparative analysis of different parts of ceramic samples. We aim to answer the research question is: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms? Our hypothesis is that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils. While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues. This innovative perspective sheds light on food practices from a microbial perspective that has not been fully discussed in archaeology.

# Background

@@

Microbial DNA extraction for 20 samples (Power soil toolkit)
Focus on bacterial, archaeal and fungal communities
Primer selection for target regions (100-300 bp): 
16S rRNA gene (V1, V4, V6) 
18S rRNA gene (ITS1)
Amplification with PCR
Gel running and collecting the bands of regions
Illumina amplicon sequencing

>>>>

# Introduction

Ceramics are the most common artifacts unearthed from archaeological sites, offering valuable insights into diverse anthropological facets encompassing cultural, social, and economic dimensions. Foodway is one of the topics. Food residues absorbed into the matrices of ceramics are useful sources to explore pottery use and past culinary practices. This analytical pursuit heavily relies on contemporary scientific methodologies. For example, organic residues, particularly lipids, extracted from pottery through a series of chemical techniques enable the identification of terrestrial animals, marine resources, and plants [@Evershed2008; @Evershed2002; @Evershed2008marine; @Craig2011; @Heron2016]. Furthermore, the identification of foodstuffs contained in storage vessels can be achieved through ancient DNA analysis, providing direct evidence of pottery content at the genes or species levels [@Foley2012; @Hansson2008; @Robinson2017]. These analytical approaches underscore the significance of preserving the condition of food residues within archaeological contexts, ensuring the efficacy of these scientific methodologies.

@@

In this study, our objective is to investigate the feasibility of using microbial communities as indicators of past food processing in pottery by focusing on the potsherds from the Guijiabao site in southwest China. As a long-term occupation site from Neolithic period to historical period, Guijiabao revealed abundant archaeological assemblages and remaines to explore the adaptation of farming communities in a mountain setting. The post-excavation analysis of Guijiabao is still in progress. This study serves as a pilot study to examine the pottery use from a perspective of microbial archaeology. Through a comparative analysis of different parts of ceramic samples and soil samples, we aim to answer the research question: Can microbial compositions serve as evidence for distinguishing various food categories based on the prevalence of spoilage microorganisms? Our hypothesis is that the microbial composition in absorbed residues within the pottery matrix will exhibit distinctions relevant to food residues when compared to those on the surface of pottery and in surrounding soils. While microorganisms offer indirect evidence of past food practices, our focus is on their potential to provide valuable information that complements existing scientific approaches to analyzing food residues. This innovative perspective sheds light on food practices from a microbial perspective that has not been fully discussed in archaeology.

# Site Description and Sampling

Guijiabao is located in the Yanyuan Basin in the middle Hengduan Mountains of southwest China. The chronological framework of Guijiabao includes the Neolithic period (5000-3700 BP), the Bronze Age (3200-2700 BP), and the historical period spanning from the late North Song to early Ming dynasties (1200-500 BP) based on radiocarbon dates [@Hao2022; @Huan2022]. Given its location along the route of population movements, Guijiabao has been a place for cultural exchanges with neighboring regions, including the northern and southern parts of the Hengduan Mountains, the Sichuan Basin, and western Yunnan since the Neolithic period. This culture interactions is evident in the similarities observed in pottery styles and decorations [@Guijiabao2017a]. Excavation at Guijiabao revealed various pottery forms, including pots, jars, urns, bowls, and basins. Predominantly, the pottery types are characterized by grayish high-tempered fine ware and reddish-brown coarse ware featuring cord-marked or basket-impressed patterns [@Hao2022; @Guijiabao2017a].

As the earliest Neolithic site in the Yanyuan Basin, Guijiabao provides valuable data for exploring the lives of people who have lived in a highland environment (approximately 2400 m above sea level). Archaeobotanical evidence and phytolith analyses suggested that mixed crop agriculture, specifically rice and millet, was introduced to this region since the occupation of the site [@Huan2022; @Guijiabao2016]. Cultivated crops, such as broomcorn millet, foxtail millet, and rice were identified as the primary agricultural practices at the site [@Deng2019; @Guijiabao2016]. Hunting and fishing could be also part of local subsistence activity, according to the finding of stone hunting implements and net sinkers [@Hao2022; @Guijiabao2020a]. Examining pottery used in daily life becomes critical in shedding light on food practices, offering more evidence for understanding local subsistence via the lens of food preparation.

As a preliminary investigation of the feasibility of using a microbial approach for detecting evidence of cooking and food types, we sampled five potsherds of typical ceramics from Guijiabao, with a pottery sample from Gaoshan, anorther Neolithic site in southwest China, as a comparison. Gaoshan is located in southwest Chengdu Plain, an alluvial plain in Sichuan, and dates back to the early Baodun period, the earliest Neolithic culture (ca. 4700-3700 BP) in the Chengdu Plain [@Gaoshan2017]. Archaeological evidence from the Gaoshan site suggests a primary subsistence pattern involving rice cultivation and pig husbandry, supplemented by millet cultivation [@Gaoshan2017; @JiangGaoshan2017; @Lee2020; @He2020]. In addition to the pottery samples, we collected three soil samples from cultural layers across various excavation units at Guijiabao, alongside two soil samples from Gaoshan.
After sampling, the potsherds were securely stored in sealed boxes, while soil samples were preserved in sterile vials, both of which were subsequently frozen at -80 °C in preparation for DNA analyses.

@@

To minimize potential sources of contamination from the environments, we cleaned all lab benches and pipettors using 99% ethanol and UV light before sample preparation and DNA extraction. These procedures were conducted in the biomolecular laboratory facilities at Academia Sinica, Taiwan. Because contemporary contamination could be a significant concern when interpreting the data, we used blank controls during sample preparation, DNA extraction, and DNA amplification to monitor extraneous DNA from surroundings, solutions and reagents, and consumables. A total of 12 powdered pottery samples (interior and exterior parts), five soil samples from Guijiabao and Gaoshan, and three negative control samples were tested in this study.

## sample preparation

Each pottery sample was divided into two distinct portions: the interior, representing the middle part of the potsherd matrix, and the exterior surface exposed to the environment. This division aimed to facilitate a comparative analysis of the microbial communities in residues absorbed into the potsherds versus those on the surface. The collection of approximately 1g of the interior potsherd was executed using a sterile drill operated at a low speed to prevent heat generation. The powdered sample was initially gathered on a pre-combustion foil and subsequently transferred into pre-sterilized tubes. A parallel sampling procedure was conducted for the surface of the potsherds, with the drill being cleaned between each iteration to minimize cross-contamination. For the soil samples, around 1g was collected from the tube containing the soil samples using sterile spatulas.

## DNA extraction and amplification

Total DNA was extracted for each sample using the DNeasy Power Soil Pro Kit (QIAGEN). All DNAs were quantified using Qubit Fluorometer.

Focus on bacterial, archaeal and fungal communities
Primer selection for target regions (100-300 bp): 
16S rRNA gene (V1, V4, V6) 
18S rRNA gene (ITS1)
Amplification with PCR
Gel running and collecting the bands of regions
Illumina amplicon sequencing
